### 2018-03-13 pubs/itp-2018/sections/metis-language-and-proofs.tex
Complete first revision to section about Metis

\section{Metis: Language and Proofs}
\label{sec:metis-language-and-proofs}

@@

\begin{figure}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{axiom~$φ₁,\cdots,φₙ$}
    \UnaryInfC{$Γ ⊢ φ₁ ∨ \cdots ∨ φₙ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{assume $φ$}
    \UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
  \end{bprooftree}
  \end{equation*}
  \vskip2mm
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ℓ ∨ φ$}
    \AxiomC{$Γ ⊢ ¬ ℓ ∨ ψ$}
    \RightLabel{resolve $ℓ$}
    \BinaryInfC{$Γ ⊢ φ ∨ ψ$}
  \end{bprooftree}
\end{equation*}
\caption{Propositional logic inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

@@

The \TPTP language is the input language to encode problems used by \Metis.
It includes the first-order form (denoted by \name{fof}) and clause normal
form (denoted by \name{cnf}) formats~\cite{sutcliffe2009}.
The \TPTP syntax\footnote{See the complete syntax grammar
at \url{http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html}} describes a
well-defined grammar to handle annotated formulas with the following form:

\begin{verbatim}
  language(name, role, formula).,
\end{verbatim}

where the \name{language} can be \name{fof} or \name{cnf}. The
\name{name} serves to identify the formula within the problem. Each
formula assumes a \name{role}, this could be an \name{axiom},
\name{conjecture}, \name{definition}, \name{plain} or an
\name{hypothesis}.

The formulas include the constants
\verb!$true! and \verb!$false!, the negation unary
operator~(\verb!~!), and the binary connectives
(\verb!&!, \verb!|!, \verb!=>!) to represent
($⊤$, $⊥$, $¬$, $∧$, $∨$, $⇒$) respectively.

@@

The \TSTP language is an output language for derivations of
ATPs~\cite{Sutcliffe-Schulz-Claessen-VanGelder-2006}.
A \TSTP derivation is a directed acyclic graph, a proof tree,
where each leaf is a formula from the \TPTP input. A node is a formula
inferred from the parent formulas. The root is the final derived formula,
such a derivation is a list of annotated formulas with the following form:

@@

The \name{source} field is an inference record with the following
pattern:

@@

The \name{rule} in the line above stands for the inference name;
the other fields are supporting arguments or useful information to
apply the reasoning step, and list the parents nodes.

\begin{notation}

We adopt a customized \TSTP syntax to keep as short as possible the
\Metis derivations for increasing the readability of this paper.

@@

Let us consider the following \TSTP derivation using
the customized \TSTP syntax (see the original \TSTP derivation and
customizations in Appendix~\ref{app:tstp-syntax}).

\begin{verbatim}
  fof(premise, axiom, p).
  fof(goal, conjecture, p).
  fof(s₀, inf(strip, goal)).
  fof(neg₀, ¬ p, inf(negate, s₀)).
  fof(n₀, ¬ p, inf(canonicalize, neg₀)).
  fof(n₁, p, inf(canonicalize, premise)).
  fof(n₂, ⊥, inf(simplify, [n₀, n₁]))
  cnf(r₀, ⊥, inf(canonicalize, n₂)).
\end{verbatim}

\end{myexamplenum}
\end{notation}

@@

\subsection{Derivations}
\label{ssec:metis-proofs}

A derivation generated by \Metis encodes a natural
deduction proof. With the inference rules in
Fig.~\ref{tab:agda-metis-table}
as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (\ie,
\emph{falsium} in the root of the \TSTP derivation).

These derivations are directed acyclic graphs, trees of refutations.
Each node stands for an application of an inference rule and the
leaves in the tree represent formulas in the given problem. Each node
is labeled with a name of the inference rule (\eg, \canonicalize).
Each edge links a premise with one conclusion. The proof graphs have
at their root the conclusion $⊥$, since \Metis derivations are
refutations.

@@

We first present the \strip inference rule since it is the rule that appears
first after each conjecture. The other rules are sorted mainly follow their
level of complexity of their definitions and the formalization presented in
Section~\ref{ssec:emulating-inferences}. Some inference rules depend on the
formalizations of other rules. For instance, the \simplify rule and the
\clausify rule need theorems developed for the \canonicalize rule. The
\canonicalize rule needs theorems developed in the \resolve formalization,
and \resolve depends on the \conjunct rule.

>>>>

\section{Metis Prover}
\label{sec:metis-language-and-proofs}

@@

\begin{figure}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{axiom~$φ₁,\cdots,φₙ$}
    \UnaryInfC{$Γ ⊢ φ₁ ∨ \cdots ∨ φₙ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{assume $φ$}
    \UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
  \end{bprooftree}
  \end{equation*}
  \vskip2mm
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ℓ ∨ φ$}
    \AxiomC{$Γ ⊢ ¬ ℓ ∨ ψ$}
    \RightLabel{resolve $ℓ$}
    \BinaryInfC{$Γ ⊢ φ ∨ ψ$}
  \end{bprooftree}
\end{equation*}
\caption{\Metis inference rules for propositional logic.}
\label{fig:metis-inferences}
\end{figure}

@@

The \TPTP language is the input language to encode problems used by \Metis.
It includes the first-order form (denoted by \name{fof}) and clause normal
form (denoted by \name{cnf}) formats~\cite{sutcliffe2009}.
The \TPTP syntax\footnote{See the complete syntax grammar
at \url{http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html}} is a
well-defined grammar to handle annotated formulas with the form:
\begin{verbatim}
  language(name, role, formula).
\end{verbatim}
where the \name{language} can be \name{fof} or \name{cnf} and the \name{name} serves
to identify the formula within the problem. Each
annotated formula also assumes a \name{role}. Some roles frequently used are \name{axiom},
\name{conjecture}, \name{definition}, \name{plain} and \name{hypothesis} but the reader
can find the full list in~\cite{sutcliffe2009}.

On the other hand, the well-written formulas can include 
the constants \verb!$true! and \verb!$false!, the negation unary
operator~(\verb!~!), and the binary connectives
(\verb!&!, \verb!|!, \verb!=>!) to represent
($⊤$, $⊥$, $¬$, $∧$, $∨$, $⇒$) respectively.
Finally, to refer us to propositional variables,
we can just write a word using letters and numbers.
%http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html#atomic_word

@@

The \TSTP language is an output language for derivations used by 
\Metis and other ATPs~\cite{Sutcliffe-Schulz-Claessen-VanGelder-2006}.
A \TSTP  derivation generated by \Metis encodes a natural
deduction proof using only as valid deduction steps the inference rules
in Fig.~\ref{tab:agda-metis-table}. 
These derivations are directed acyclic graphs, trees of refutations.
Each node stands for an application of an inference rule to the formulas
in the parent nodes. The leaves in the tree represent formulas in the given
\TPTP input. Each node is labeled with a name of its inference rule (\eg, \canonicalize).
Each edge in the tree links a premise with one conclusion.

Since \Metis attempts to prove conjectures by refutation, and the root
is the final derived formula in these proof trees,  we have at their
roots the conclusion $⊥$. Such derivations are a list of annotated
formulas with the form:

@@

The new field regarding the \TPTP syntax is the \name{source} field
that mainly intends to record the inference rule of the deduction step.
This field has the following pattern:

@@

We have the \name{rule} field with the inference name, another field
including supporting arguments to apply the reasoning step, and the
last one field with a list its parents nodes.

For increasing the readability of this paper, we are not presenting
these \TSTP derivations using the above description. Instead of it, we
have adopted a customized \TSTP syntax to keep as short as possible
the \Metis derivations. For instance,  we shortened some keywords,
removed unnecessary arguments, and replaced some symbols. The list of
all customizations is in Appendix~\ref{app:tstp-syntax}.

@@

Let us consider the following \TSTP derivation generated by \Metis after using
the customized \TSTP syntax proposed for the theorem:

\begin{equation}
(p \Rightarrow q) \wedge (q \Rightarrow p) ⊢ (p \vee q) \Rightarrow (p \wedge q).
\end{equation}

% this tstp example should be used to explain the reconstruction.
\begin{verbatim}
  fof(premise, axiom, (p ⇒ q) ∧ (q ⇒ p)).
  fof(goal, conjecture, (p ∨ q) ⇒ (p ∧ q)).
  fof(s₀, (p ∨ q) ⇒ p, inf(strip, goal)).
  fof(s₁, ((p ∨ q) ∧ p) ⇒ q, inf(strip, goal)).
  fof(neg₀, ¬ ((p ∨ q) ⇒ p), inf(negate, s₀)).
  fof(n₀₀, (¬ p ∨ q) ∧ (¬ q ∨ p), inf(canonicalize, premise)).
  fof(n₀₁, ¬ q ∨ p, inf(conjunct, n₀₀)).
  fof(n₀₂, ¬ p ∧ (p ∨ q), inf(canonicalize, neg₀)).
  fof(n₀₃, p ∨ q, inf(conjunct, n₀₂)).
  fof(n₀₄, ¬ p, inf(conjunct, n₀₂)).
  fof(n₀₅, q, inf(simplify, [n₀₃, n₀₄])).
  cnf(r₀₀, ¬ q ∨ p, inf(canonicalize, n₀₁)).
  cnf(r₀₁, q, inf(canonicalize, n₀₅)).
  cnf(r₀₂, p, inf(resolve, q, [r₀₁, r₀₀])).
  cnf(r₀₃, ¬ p, inf(canonicalize, n₀₄)).
  cnf(r₀₄, ⊥, inf(resolve, p, [r₀₂, r₀₃])).
  fof(neg₁, ¬ ((p ∨ q) ∧ p) ⇒ q), inf(negate, s₁)).
  fof(n₁₀, ¬ q ∧ p ∧ (p ∨ q), inf(canonicalize, neg₁)).
  fof(n₁₁, (¬ p ∨ q) ∧ (¬ q ∨ p), inf(canonicalize, premise)).
  fof(n₁₂, ¬ p ∨ q, inf(conjunct, n₁₁)).
  fof(n₁₃, ⊥, inf(simplify,[n₁₀, n₁₂])).
  cnf(r₁₀, ⊥, inf(canonicalize, n₁₃)).
\end{verbatim}

\end{myexamplenum}

@@

@@

We first present the reconstruction of the \strip inference rule and its
description since it is the rule that appears
first after each conjecture in \TSTP derivations. The other rules are sorted mainly follow their
level of complexity of their definitions and the formalization presented in
Section~\ref{ssec:emulating-inferences}. Some inference rules depend on the
formalizations of other rules. For instance, the \simplify rule and the
\clausify rule need theorems developed for the \canonicalize rule. The
\canonicalize rule needs theorems developed in the \resolve formalization,
and \resolve depends on the \conjunct rule.
